---
title: "Covid Flights Analysis"
author: "Group 2"
date: "5/6/2022"
output: html_document
---
```{r, global_options, include = FALSE}
knitr::opts_chunk$set(message = FALSE, tidy.opts = list(width.cutoff = 60), tidy = TRUE)
knitr::opts_chunk$set(echo = TRUE)
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#put all libraries used in the report here 
library(tidyverse)
library(patchwork)
library(ggplot2)
library(tsibble)
library(tsibbledata)
library(lubridate)
library(feasts)
library(readr)
library(ggridges)
```

# Guiding Question
How did significant factors affect airline flight delays before and during COVID-19? 

# Introduction and Overview of Data
  This study is exploring factors that contributed to arrival delays in U.S. airline traffic during the beginning of the COVID-19 pandemic. Two primary data sets were used. The first data set was found on Kaggle, but was collected from the U.S. Department of Transportation's (DOT) Bureau of Transportation Statistics (Transportation, 2020); and contains information related to airline flights, flight date, arrival delays, destination state, air time, distance, and flight carrier. The second data set is from an ongoing repository of data on coronavirus cases and deaths in the U.S., from the New York Times (Coronavirus, 2022); and contains information about COVID-19 cases and deaths by state since the start of the pandemic. 
  The aim is to look into how flight arrival delay was impacted by the COVID-19 pandemic. Arrival delay is compared with origin of flight, airline carrier, and number of cases per day per state. Arrival delay is analyzed pre- and post-COVID, of which the start of the pandemic is deemed to be March 1, 2020 (this date chosen because of the dramatic increase in cases during this month).


# Data Processing 
 	The U.S. flights data set contained 47 variables. Out of those variables, 19 were explored based on what was deemed relevant to the guiding question. We also joined the Flights data set with a COVID data set (by date and state [of departure]) to get access to two more variables, cases and deaths, for a total of 21 variables. 
	Many observations contained “NA” values (roughly 89% of the data) and the “canceled” variable was causing this feature, as the other delay statistics could not pertain to a flight that had never taken off. As these “NA” features were logical, the corresponding rows were kept in the data set. However, there were some “NA” values in the Flights data set and the corresponding observations were removed (~0.2% of the data).
	
```{r data processing, cache = TRUE, message = FALSE}
flights <- read_csv("./jantojun2020.csv")
us_states <- read_csv("https://raw.githubusercontent.com/nytimes/covid-19-data/master/us-states.csv")

desired_cols <- c(2,3,4,5,6,7,10,13,14,17,18,20,25,28,29,31,36,40,41) # desired columns

# selecting columns, correcting column data types, and eliminating incomplete rows
clean_flights0 <- flights %>%
  select(desired_cols) %>%
  mutate(DISTANCE = as.double(DISTANCE),
         CANCELLED = as.logical(CANCELLED),
         FL_DATE = as.Date(FL_DATE,format = "%m/%d/%y"))

us_states2 <- us_states [,-3] %>% # removing "fips" column and dates after flights dataset
  mutate(date = as.Date(date)) %>%
  subset(date <= max(clean_flights0$FL_DATE)) # dropping data after the latest flight date

clean_flights1 <- clean_flights0 %>%
  slice(-which(rowSums(is.na(clean_flights0)) > 0 & 
                 clean_flights0$CANCELLED == 0))
# continuous values
c_clean <- clean_flights1 %>%
  select_if(is.double)
# discrete values
d_clean1 <- clean_flights1 %>% 
  select_if(Negate(is.double))

# discrete values, followed by continuous
new_flights <- cbind(d_clean1, c_clean)

# left join by date and ORIGIN STATE
fulldata <- left_join(new_flights, us_states2, by = c(c("FL_DATE" = "date"), c("ORIGIN_STATE_NM" = "state")))

fulldata3 <- fulldata %>%
  replace(is.na(.), 0) %>% # converting NA values to 0
  mutate(cases = as.integer(cases), # making discrete column integer
         deaths = as.integer(deaths)) # making discrete column integer
```

# Analysis of Key Variables
```{r random sample, include = FALSE}
slim_data = fulldata3 %>% 
  select(ARR_DELAY, DISTANCE, ORIGIN, ORIGIN_STATE_NM, deaths)
names(slim_data) = tolower(names(slim_data))

#speed up initial plot development by randomly sampling 10% of the population
sample_row_index = sample(1:nrow(slim_data), nrow(slim_data)/10)

#so i can change the col names to lowercase to work with code I have written without messing up others code
fulldata3_copy = fulldata3
names(fulldata3_copy) = tolower(names(fulldata3_copy))

#slim_sample is now the full data so plots represent all data
slim_sample = fulldata3_copy #slim_data[sample_row_index,] 
```
```{r random sample, include = FALSE}
slim_data = fulldata3 %>% 
  dplyr::select(ARR_DELAY, DISTANCE, ORIGIN, ORIGIN_STATE_NM, deaths)
names(slim_data) = tolower(names(slim_data))

#speed up initial plot development by randomly sampling 10% of the population
sample_row_index = sample(1:nrow(slim_data), nrow(slim_data)/10)

#so i can change the col names to lowercase to work with code I have written without messing up others code
fulldata3_copy = fulldata3
names(fulldata3_copy) = tolower(names(fulldata3_copy))

#slim_sample is now the full data so plots represent all data
slim_sample = fulldata3_copy #slim_data[sample_row_index,]  
```

## Arrival Delay and Origin
  One variable that seems to significantly influence arrival delays is the airport the flight originated from. A one way ANOVA test of delay modeled by origin reveals that there are statistically significant differences in arrival delays among airports (p-value = $1*10^{-163}$. The rejection of the null hypothesis is unsurprising as there are 375 airports in the data set. The best and the worst airports to fly from during the covid pandemic are shown below.
```{r anova origin, cache = TRUE}
# The full data set is too large to run ANOVA on,
# but running it on a randomly sampled subset reveals that the null is rejected with a pvalue of essentially 0.
sample_row_index = sample(1:nrow(slim_data), nrow(slim_data)/10)
slim_sample = slim_data[sample_row_index,] 
model = aov(arr_delay ~ origin, data = slim_sample)
results = anova(model)
```

```{r arrival delay and origin, cache = TRUE, fig.height = 5, fig.width = 10}
longest_origin_delays = slim_sample %>% 
  group_by(origin) %>% 
  summarize(mean_delay = mean(arr_delay)) %>% 
  arrange(desc(mean_delay)) %>% 
  head(10)

shortest_origin_delays = slim_sample %>% 
  group_by(origin) %>% 
  summarize(mean_delay = mean(arr_delay)) %>% 
  arrange(desc(mean_delay)) %>% 
  tail(10)

p1 = ggplot(longest_origin_delays) + 
  geom_col(aes(mean_delay, rev(reorder(origin, mean_delay)))) +
  labs(x = "Mean Delay (min)", 
       y = "Origin",
       title = "10 Worst Airports to Fly From") +
  scale_y_discrete(labels = c("Sheridan, WY",
                              "Lewiston, ID",
                              "W. Regional, NE",
                              "Hattiesburg, MI",
                              "Cold Bay, AL",
                              "Yellowstone, WY",
                              "Mobile Bay, AL",
                              "Cheyenne, WY",
                              "Hobs, NM",
                              "Saywer, MI"))

p2 = ggplot(shortest_origin_delays) + 
  geom_col(aes(abs(mean_delay), reorder(origin, mean_delay),)) +
  labs(x = "Mean Minutes Arrived Early", 
       y = "",
       title = "10 Best Airports to Fly From") +
  scale_y_discrete(labels = c("Lib. Regional, KS",
                              "Decatur, IL",
                              "Yakutat, AK",
                              "Int. Falls, MN",
                              "Pierre, SD",
                              "Houghton, MI",
                              "Martha's Vineyard, MA", 
                              "Quincy, IL", 
                              "W. Yellowstone", 
                              "Pudacah, KY"))
 
p = p1 + p2
p
#ggsave("airport.jpeg", p, width = 10, height = 7, units = "in")
```

## Arrival Delay and Airline Carrier
  There was some variation in mean delay per day for the different airline carriers. A one-way ANOVA test showed there were significant differences in delay time for the different airline carriers ($p<<0.05$). Eliminating points greater than 3 standard deviations away from the mean delay for all carriers, differences across carriers were easily observable. Again, a one-way ANOVA test garners the same result, though it is less statistically significant ($p=0.02$). 
  
```{r formatting data for time series, cache=TRUE, echo=FALSE}
slim_data2 <- fulldata3 %>% 
  dplyr::select(ARR_DELAY, FL_DATE, DEST_STATE_NM, cases, MKT_UNIQUE_CARRIER)
names(slim_data2) = tolower(names(slim_data2)) 
slim_data2 <- slim_data2 %>%
  mutate(pandemic = ifelse(fl_date < "2020-03-01", "pre", "during"))

group_cols <- c("fl_date", "dest_state_nm")
#group data by date and destination, and summarize mean delays and cases
group_state <- slim_data2 %>%
  group_by(across(all_of(group_cols))) %>%
  summarise(delay_mean_on_given_day = mean(arr_delay),
            cases_mean_on_given_day = mean(cases)) 
group_state <- as.data.frame(group_state) #ungroup for sibble
#tsibble indexed by date 
group_state_ts <- as_tsibble(group_state, key = c("dest_state_nm"))
group_state_ts <- group_state_ts %>%
  mutate(pandemic = ifelse(fl_date < "2020-03-01", "pre", "during"))
#group data by date and carrier, and summarize mean delays and cases
group_carrier <- slim_data2 %>%
  group_by(across(all_of(c("fl_date", "mkt_unique_carrier")))) %>%
  summarise(delay_mean_on_given_day = mean(arr_delay),
            cases_mean_on_given_day = mean(cases)) 
group_carrier <- as.data.frame(group_carrier) #ungroup for sibble
#tsibble indexed by date 
group_carrier_ts <- as_tsibble(group_carrier, key = c("mkt_unique_carrier"))

#filter to midwest destination states and dates after March that will have more cases
midwest <- c("Illinois", "Indiana", "Iowa", "Kansas", "Michigan", "Minnesota", "Missouri",
             "Nebraska", "South Dakota", "Ohio", "North Dakota", "Wisconsin")
western <- c("Alaska", "Arizona", "California", "Colorado", "Hawaii", "Idaho", "Montana",
             "Nevada", "New Mexico", "Oregon", "Utah", "Washington", "Wyoming")
midwest_df <- group_state_ts[group_state_ts$dest_state_nm %in% midwest, ]
western_df <- group_state_ts[group_state_ts$dest_state_nm %in% western, ]
```

```{r arrival delay by airline carrier, cache=TRUE, fig.width=16, fig.height=10}
group_carrier_ts1 <- group_carrier_ts %>% 
  mutate(across(mkt_unique_carrier, ~ recode(., "AA" = "American Airlines", "AS" = "Alaska Airlines", "B6" = "JetBlue",
                                             "DL" = "Delta Air Lines", "F9" = "Frontier Airlines", "G4" = "Allegiant Air",
                                             "HA" = "Hawaiin Airlines", "NK" = "Spirit Airlines", "UA" = "United Airlines",
                                             "WN" = "Southwest Airlines")))

p1 <- ggplot(group_carrier_ts1) +
  geom_boxplot(aes(y=delay_mean_on_given_day, x=reorder(mkt_unique_carrier,delay_mean_on_given_day, mean),
               fill=mkt_unique_carrier)) +
  geom_hline(yintercept = mean(group_carrier_ts$delay_mean_on_given_day), alpha=0.5, col="red", size=1) +
  scale_fill_brewer(palette="Set3") +
  theme(legend.position = "none",
        axis.title = element_text(size=12),
        axis.text.x = element_text(angle = 45, hjust=1)) +
  labs(y="Mean delay time per day", x="Airline carrier", title="Mean delay per day by carrier")

#test for differences in carrier delay times
#summary(aov(delay_mean_on_given_day ~ mkt_unique_carrier, data=group_carrier_ts1))

#test for differences, removing delays more than 3 standard deviations away from the mean delay time
sds <- mean(group_carrier_ts1$delay_mean_on_given_day) + (c(-3,3) * sd(group_carrier_ts1$delay_mean_on_given_day))
outlier.omit <- group_carrier_ts1 %>% 
  filter(delay_mean_on_given_day<sds[1] | delay_mean_on_given_day>sds[2])

p2 <- ggplot(outlier.omit) +
  geom_boxplot(aes(y=delay_mean_on_given_day, x=reorder(mkt_unique_carrier,delay_mean_on_given_day, mean),
               fill=mkt_unique_carrier)) +
  geom_hline(yintercept = mean(group_carrier_ts$delay_mean_on_given_day), alpha=0.5, col="red", size=1) +
  scale_fill_brewer(palette="Set3") +
  theme(legend.position = "none",
        axis.title = element_text(size=12),
        axis.text.x = element_text(angle = 45, hjust=1)) +
  labs(y="Mean delay time per day", x="Airline carrier", title="Mean delay per day by carrier (delays > 3 SD from mean removed)")

#summary(aov(delay_mean_on_given_day ~ mkt_unique_carrier, data=outlier.omit))

p1 | p2
```

## Arrival Delays and Flights During the COVID-19 Pandemic

  Arrival delays were compared with COVID deaths on that particular day, serving as a proxy for the severity of COVID. Even though these data are from near the peak of COVID, only 42% of the dates in this range have any COVID deaths at all. The Spearman's correlation coefficient for arrival delays and deaths is -0.05. A graphical approach with the scatter plot also suggests a weak association.
  
```{r arrival delay and deaths, cache = TRUE, echo=FALSE}
death_correlation = cor(slim_sample$arr_delay, slim_sample$deaths, method = "spearman")
zero_death_days = sum(slim_sample$deaths == 0) / length(slim_sample$deaths)
ggplot(slim_sample, aes(deaths, arr_delay)) + 
  geom_point() +
  labs(x = "Covid Related Deaths",
       y = "Arrival Delay",
       title = "Covid Deaths vs Arrival Delay")
```

  There is an initial increase in mean delay times directly preceding and during the first month of the COVID-19 pandemic when cases began exponentially increasing. This makes sense given that domestic flight departures were heavily reduced by as much as 71.5% in May 2020 compared to May 2019 (1). After the initial increase, overall delays appear to decrease and exhibit significantly less variation. This could in part be attributed to there being a lower number of domestic flights. 
  

```{r Midwest and Western US time series, cache=TRUE, fig.height=6, fig.width=11, echo=FALSE}
#mean cases per day by midwest destination state (after March 2020)
p1 <- ggplot(midwest_df%>%filter(fl_date>="2020-03-01")) +
        geom_line(aes(x = fl_date, y = cases_mean_on_given_day, col=dest_state_nm)) +
        scale_fill_hue(l=100) +
        guides(color = guide_legend(override.aes = list(size = 3))) +
        labs(y="Mean cases per day (100,000s)", x="", title = "Mean cases per day (midwest)") +
        scale_y_continuous(label = scales::label_number(scale=1/1e5)) +
        scale_color_brewer(palette="Paired", name="destination state") +
       # scale_color_discrete(name = "destination state") +
        scale_x_date(expand=c(0,0)) +
        theme(axis.title = element_text(size=10),
              legend.position = "bottom")
#mean delay time per day by selected midwest destination state
p2 <- ggplot(midwest_df %>% filter(dest_state_nm==c("Wisconsin","Illinois"))) +
        geom_line(aes(x = fl_date, y = delay_mean_on_given_day, col=dest_state_nm), size=1) +
        scale_fill_hue(l=100) +
        guides(color = guide_legend(override.aes = list(size = 3))) +
        labs(title="Mean delay time per day (WI & IL)", y="Mean delay time per day (minutes)", x="") +
        scale_color_manual(values = c("#A6CEE3", "#B15928"), name = "") +
        scale_x_date(expand=c(0,0)) +
        ylim(-20,100) +
        theme(legend.position = "bottom",
              axis.title = element_text(size=10))

#mean cases per day by western destination state
p3 <- ggplot(western_df %>% filter(fl_date>="2020-03-01")) +
        geom_line(aes(x = fl_date, y = cases_mean_on_given_day, col=dest_state_nm)) +
        scale_fill_hue(l=100) +
        guides(color = guide_legend(override.aes = list(size = 3))) +
        labs(y="Mean cases per day (in 100,000s)", x="", title = "Mean cases per day (western)") +
        scale_y_continuous(label = scales::label_number(scale=1/1e5)) +
        scale_color_discrete(name = "") +
        scale_x_date(expand=c(0,0)) +
        theme(axis.title = element_text(size=10),
              legend.position = "bottom")
#mean delay time per day by western destination state 
p4 <- ggplot(western_df %>% filter(dest_state_nm=="California")) +
        geom_line(aes(x = fl_date, y = delay_mean_on_given_day, col=dest_state_nm)) +
        scale_fill_hue(l=100) +
        guides(color = guide_legend(override.aes = list(size = 3))) +
        labs(y="Mean delay time per day (minutes)", x="") +
        scale_x_date(expand=c(0,0)) +
        ylim(-20,100) +
        theme(legend.position = "none",
              axis.title = element_text(size=10))


p1 + p3
p2
```

  Globally, the number of flights decreased by 50% in April and May 2020 (2). 8 of the 10 days with the most flights—all in March 2020—coincided with the surge of COVID-19 cases.
  
```{r number of flights time series, cache = TRUE, fig.height = 5, fig.width = 8, echo=FALSE}
#count number of flights per day for each destination state
number_flights <- slim_data2 %>%
  group_by(dest_state_nm) %>%
  count(fl_date, sort=TRUE)

most_flights <- number_flights[0:10,] #CA has 10 of the top 10 number of flight days by state

number_flights_ts <- slim_data2 %>%
  group_by(dest_state_nm) %>%
  count(fl_date, sort=TRUE) %>%
  as.data.frame() %>%
  as_tsibble(key="dest_state_nm")

#just looking at western states again since California has the most flights
western_no_df <- number_flights_ts[number_flights_ts$dest_state_nm %in% western, ]

ggplot(western_no_df) +
  geom_line(aes(x = fl_date, y = n, col=dest_state_nm)) +
  scale_fill_hue(l=100) +
  guides(color = guide_legend(override.aes = list(size = 3))) +
  labs(y="Number of flights", x="", title="Number of flights in Western States") +
  scale_x_date(expand=c(0,0)) +
  scale_color_discrete(name = "destination state") +
  theme(legend.position = "bottom",
        axis.title = element_text(size=10))
```

   The average delay time per day for pre-pandemic flights appears to be higher than flights during the pandemic, which is the case for nearly every state. There is also greater variation in delay time for pre-pandemic flights.

```{r, fig.height=8, fig.width=18, cache=TRUE, warning=FALSE, message=FALSE, echo=FALSE}
#ridge plots for mean delays by state
p_west <- ggplot(western_df, 
                 aes(x = `delay_mean_on_given_day`, y = reorder(`dest_state_nm`, delay_mean_on_given_day, sd), 
                     fill = pandemic)) +
  geom_density_ridges(alpha=.8, scale = 2, rel_min_height = 0.01) +
  scale_x_continuous(expand = c(0, 0)) +
  scale_y_discrete(expand = c(0, 0)) +
  coord_cartesian(clip = "off") +
  labs(title = 'Mean delays by state (western)', x="mean delay per day") +
  theme_ridges(font_size = 13, grid = TRUE) +
  theme(axis.title.y = element_blank(),
        legend.position = "bottom",
        axis.title.x = element_text(hjust=0.5)) +
  scale_fill_brewer(palette="RdPu", direction=-1)

p_mid <- ggplot(midwest_df,
                aes(x = `delay_mean_on_given_day`, y = reorder(`dest_state_nm`, delay_mean_on_given_day, sd),
                    fill = pandemic)) +
  geom_density_ridges(alpha=.8, scale = 2, rel_min_height = 0.01) +
  scale_x_continuous(expand = c(0, 0)) +
  scale_y_discrete(expand = c(0, 0)) +
  coord_cartesian(clip = "off") +
  labs(title = 'Mean delays by state (midwest)', x="mean delay per day") +
  theme_ridges(font_size = 13, grid = TRUE) +
  theme(axis.title.y = element_blank(),
        legend.position = "bottom",
        axis.title.x = element_text(hjust=0.5)) +
  scale_fill_brewer(palette="Paired", direction=-1)

p_mid + p_west
```

  A Kolmogorov-Smirnov test was run to test whether the distributions of delays per day are significantly different pre-pandemic than during the pandemic. The test revealed that delays are significantly different ($p<<0.05$). Mean delay time per day pre-pandemic ($0.22 \pm 13.37$) was greater than mean delay time per day ($-7.64 \pm 9.46$) during the pandemic. 
  
```{r k-s test, cache=TRUE, evaluate=FALSE}
pre_pandemic <- group_state_ts[group_state_ts$pandemic == "pre", ]
pandemic <- group_state_ts[group_state_ts$pandemic == "during", ]

#mean(pre_pandemic$delay_mean_on_given_day) #mean delay time pre-pandemic
#sd(pre_pandemic$delay_mean_on_given_day)

#mean(pandemic$delay_mean_on_given_day) #mean delay time pandemic
#sd(pandemic$delay_mean_on_given_day)

#Kolmogorov-Smirnov test for difference in distributions of mean delays per day for pre-pandemic and pandemic flights
ks.test(pre_pandemic$delay_mean_on_given_day, pandemic$delay_mean_on_given_day)
```
# Conclusion

From our six visualizations, readers can observe the airline carriers’ performance during the COVID-19 pandemic. 
In addition, they can view the best and worst ten airports to fly from and select the better airline carrier. 
From time series analysis, the trend of the number of flights during the pandemic is shown along with the mean COVID cases per day. 
Readers can retrieve the relationship between cases per day and the number of flights. 
Although our visualization does not cover a deeper analysis of other flight data, 
it still provides readers with a solid understanding of the impact of COVID-19 on flights in the United States. 
We hope our visualization will allow readers to choose the best airline carriers and airports for their travel plans.

# Notes

1. Abate, M., Christidis P., and Purwanto, A.J. (2020). Government support to airlines in the aftermath of the COVID-19 pandemic. _Journal of Air Transport Management, 89_, 101931. https://doi.org/10.1016/j.jairtraman.2020.101931. 

2. Coronavirus in the U.S.: Latest Map and Case Count. (2022). _New York Times Company_. Retrieved from https://www.nytimes.com/interactive/2021/us/covid-cases.html. 

3. Hotle, S. and Mumbower, S. (2021). The impact of COVID-19 on domestic U.S. air travel operations and commercial airport service. _Transportation Research Interdisciplinary Perspectives, 9_, 100277. https://doi.org/10.1016/j.trip.2020.100277.

4. Transportation Statistics Annual Report 2020. (2020, December 1). _United States. Department of Transportation. Bureau of Transportation Statistics_. https://doi.org/10.21949/1520449. 

