---
title: "data_processing"
author: "Group 2"
date: "3/13/2022"
output: pdf_document
---

```{r, include = FALSE, echo = FALSE, message = FALSE, warning = FALSE}
# Column Descriptions

# COVID-19 Airline Flight Delays
# 1:YEAR (Year)
# 2:QUARTER (Jan-Mar, 2: Apr-Jun, 3: Jul-Sep, 4: Oct-Dec)
# 3:MONTH (Month of Year)
# 4:DAY_OF_MONTH (Date of Month)
# 5:DAY_OF_WEEK (Day of Week [1: Monday, 7: Sunday])
# 6:FL_DATE (Full flight date [M/DD/YYYY])
# 7:MKT_UNIQUE_CARRIER (AA: American Airlines AS: Alaska Airlines B6: JetBlue DL: Delta Air Lines F9: Frontier Airlines G4: Allegiant Air HA: Hawaiian Airlines NK: Spirit Airlines UA: United Airlines WN: Southwest Airlines)
# 8:MKT_CARRIER_FL_NUM (Flight Number)
# 9:TAIL_NUM (Aircraft Tail Number [Usually starts with 'N'])
# 10:ORIGIN (Flight Departure 3-Letter Airport Abbreviation)
# 11:ORIGIN_CITY_NAME (Flight Departure City Name)
# 12:ORIGIN_STATE_ABR (Flight Departure 2-Letter State Abbreviation)
# 13:ORIGIN_STATE_NM (Flight Departure City, State Abbreviation)
# 14:DEST (Flight Arrival 3-Letter Airport Abbreviation)
# 15:DEST_CITY_NAME (Flight Arrival City, State Abbreviation)
# 16:DEST_STATE_ABR (Flight Arrival 2-Letter State Abbreviation)
# 17:DEST_STATE_NM (Flight Arrival State Name)
# 18:CRS_DEP_TIME (Scheduled Departure Time [HHMM] [Single or 2-Digit Values Represent 00:MM, e.g. 3 represents 00:03 or 12:03 AM])
# 19:DEP_TIME (Actual Departure Time [HHMM])
# 20:DEP_DELAY (Departure Delay [Difference Between Actual Departure Time and Scheduled Departure Time in Minutes])
# 21:DEP_DELAY_NEW (Departure Delay Ignoring Early Departures [Listed as 0])
# 22:DEP_DEL15 (Departure Delay Greater Than 15 Minutes [0: Not Greater Than 15, 1: Greater Than 15])
# 23:DEP_DELAY_GROUP (Departure Delay in Number of 15-minute increments Rounded Down [e.g. Early Departure (< 0) is a value of -1, 30 or 42 minutes is a value of 2])
# 24:DEP_TIME_BLK (Scheduled Departure Time in Hourly Block [HHMM])
# 25:TAXI_OUT (Time between Airplane Taxi from Gate and Takeoff [WHEELS_OFF] Time [in Minutes])
# 26:WHEELS_OFF (Time of Airplane Takeoff [HHMM])
# 27:WHEELS_ON (Time of Airplane Landing [HHMM])
# 28:TAXI_IN (Time between Airplane Taxi to Gate and Landing [WHEELS_ON] Time [in Minutes])
# 29:CRS_ARR_TIME (Scheduled Arrival Time [HHMM] [Single or 2-Digit Values Represent 00:MM, e.g. 3 represents 00:03 or 12:03 AM])
# 30:ARR_TIME (Actual Arrival Time [HHMM])
# 31:ARR_DELAY Arrival Delay (Difference Between Actual Arrival Time and Scheduled Arrival Time in Minutes)
# 32:ARR_DELAY_NEW (Arrival Delay Ignoring Early Arrivals [Listed as 0])
# 33:ARR_DEL15 (Arrival Delay Greater Than 15 Minutes [0: Not Greater Than 15, 1: Greater Than 15])
# 34:ARR_DELAY_GROUP (Arrival Delay in Number of 15-minute increments Rounded Down [e.g. Early Arrival (< 0) is a value of -1, 30 or 42 minutes is a value of 2])
# 35:ARR_TIME_BLK (Scheduled Arrival Time in Hourly Block [HHMM])
# 36:CANCELLED (0: Flight Not Cancelled, 1: Flight Cancelled)
# 37:CANCELLATION_CODE (Reason for Cancellation - if Cancelled, Letter Present [A: Carrier, B: Weather, C: National Aviation System, D: Security])
# 38:CRS_ELAPSED_TIME (Scheduled Total Flight Time [in Minutes])
# 39:ACTUAL_ELAPSED_TIME (Actual Total Elapsed Flight Time [in Minutes])
# 40:AIR_TIME (Actual Total Elapsed Time Airplane in the Air [in Minutes])
# 41:DISTANCE (Distance Between Departure and Arrival Airports [in Miles])
# 42:DISTANCE_GROUP (Distance Between Departure and Arrival Airports in Number of 250-Mile increments Rounded Down [e.g. 400 miles is a value of 1])
# 43:CARRIER_DELAY (Carrier Delay [in Minutes])
# 44:WEATHER_DELAY (Weather Delay [in Minutes])
# 45:NAS_DELAY (National Aviation System Delay [in Minutes])
# 46:SECURITY_DELAY (Security Delay [in Minutes])
# 47:LATE_AIRCRAFT_DELAY (Late Aircraft Delay [in Minutes])

# NY Times Covid 19 Data: US States
# 1:date (Date [YYYY-MM-DD])
# 2:state (State Name)
# 3:fips (State Federal Information Processing Standards Code)
# 4:cases (Number of Cases)
# 5:deaths (Number of Deaths)

```

# Part 1 - Data Processing
```{r, include = TRUE, echo = FALSE, message = FALSE, warning = FALSE}
library(dplyr)
library(tidyr)
library(ggplot2)
library(ggpubr)
library(ggridges)
library(agricolae)

flights <- read.csv("./jantojun2020.csv")
us_states <- read.csv("https://raw.githubusercontent.com/nytimes/covid-19-data/master/us-states.csv")
```

## 1.2 Filtering Data
```{r, include = TRUE, echo = FALSE, message = FALSE, warning = FALSE}
desired_cols <- c(2,3,4,5,6,7,10,13,14,17,18,20,25,28,29,31,36,40,41) # desired columns
names(flights[,desired_cols]) # columns kept
flights2 <- flights[,desired_cols] # dataframe with correct columns
```

## 1.3 Correcting Data Types
```{r, include = TRUE, echo = FALSE, message = FALSE, warning = FALSE}
clean_flights0 <- flights2 %>% # correcting column data types
  mutate(DISTANCE = as.double(DISTANCE)) %>% 
  mutate(CANCELLED = as.logical(CANCELLED)) %>%
  mutate(FL_DATE = as.Date(FL_DATE,format = "%m/%d/%y"))
```

## 1.4 Incomplete Data
```{r, include = TRUE, echo = FALSE, message = FALSE, warning = FALSE}
# CANCELLED column leaves NA in other columns
clean_flights0_length <- nrow(clean_flights0) # number of rows in dataset

# number of not cancelled flights with missing data
num_na_rows <- length(which(rowSums(is.na(clean_flights0)) > 0 & clean_flights0$CANCELLED == 0)) 
share_na <- num_na_rows / clean_flights0_length # share of rows with NA values

num_filled_rows <- clean_flights0_length - num_na_rows # number of fully filled rows

share_na # share of rows with missing information
num_filled_rows # number of rows with full information

bad_rows <- which(rowSums(is.na(clean_flights0)) > 0 & clean_flights0$CANCELLED == 0) # rows  with missing information
clean_flights1 <- clean_flights0[-bad_rows,] # eliminating incomplete rows and correcting data types
```

## 1.5 Continuous vs. Discrete Data
```{r, include = TRUE, echo = FALSE, message = FALSE, warning = FALSE}
c_clean <- clean_flights1 %>% # non-numeric dates
            select_if(is.double)

# only continuous columns w/ NA values
c_clean0 <- clean_flights0 %>% 
            select_if(is.double) %>%
            mutate(FL_DATE = as.numeric(clean_flights0$FL_DATE)) 

# only continuous columns w/o NA values
c_clean1 <- clean_flights1 %>% 
            select_if(is.double) %>%
            mutate(FL_DATE = as.numeric(clean_flights1$FL_DATE)) 

# only discrete columns w/o NA values
d_clean1 <- clean_flights1 %>% select_if(Negate(is.double))

# only continuous columns w/ NA values (randomly sampled)
sampled_rows <- sample(1:nrow(d_clean1), nrow(d_clean1), replace = FALSE)
d_clean0 <- clean_flights0[sampled_rows,]
```

## 1.6 Goodness of Fit Testing

### 1.6.1 Calculations
```{r, include = TRUE, echo = FALSE, message = FALSE, warning = FALSE}
#OPTIONAL
c_pvals <- c() # creating empty vector for p-values
for(i in 1:(ncol(c_clean1))) { # running Kolmogorov-Smirnov Test (continuous GOF test) between all columns
  ks_result <- ks.test(c_clean0[,i], c_clean1[,i]) # running K-S Test
  c_pvals <- append(c_pvals, ks_result$p) # appending p-value to p-value vector
}
c_pval_table <- matrix(c_pvals, ncol = length(c_pvals), byrow = FALSE) # creating table with p-values
colnames(c_pval_table) <- colnames(c_clean1) # matching columns with appropriate column names
c_pval_table # p-value table

#OPTIONAL
d_pvals <- c() # creating empty vector for p-values
for(i in 1:(ncol(d_clean1))) { # running Kolmogorov-Smirnov Test (continuous GOF test) between all columns
  chisq_result <- chisq.test(d_clean0[,i], d_clean1[,i]) # running Pearson's Chi-Squared Test
  d_pvals <- append(d_pvals, chisq_result$p.value) # appending p-value to p-value vector
}
d_pval_table <- matrix(d_pvals, ncol = length(d_pvals), byrow = FALSE) # creating table with p-values
colnames(d_pval_table) <- colnames(d_clean1) # matching columns with appropriate column names
d_pval_table # p-value table
```

### 1.6.2 Visualization 
```{r echo=FALSE, message=FALSE, warning=FALSE, include=TRUE}
#OPTIONAL
for (i in 1:(ncol(c_clean1))) { # histograms of all continuous columns: raw vs. processed
  print(ggplot() + 
          geom_density(aes(c_clean0[,i]), 
                       color = "black", 
                       fill = "black", 
                       alpha = 0.8) + 
          geom_density(aes(c_clean1[,i]), 
                       color = "blue", 
                       fill = "blue", 
                       alpha = 0.3) + 
          labs(title = colnames(c_clean1)[i], 
               x = "value", 
               y = "density") +
          theme(strip.text.x = element_text(size = 4), 
                axis.text.x = element_blank(), 
                axis.ticks.x=element_blank()) +
          annotate("text", -Inf, Inf, label = "Processed", hjust = 0, vjust = 1, color = "blue") + 
          annotate("text", -Inf, Inf, label = "Raw", hjust = 0, vjust = 3, color = "black"))
}

#OPTIONAL
for (i in 1:(ncol(d_clean1))) { # barplots of all discrete columns: raw vs. processed
  print(ggplot() + 
          geom_bar(aes(d_clean0[,i]), 
                   color = "black", 
                   fill = "black", 
                   alpha = 0.8) + 
          geom_bar(aes(d_clean1[,i]), 
                   color = "blue", 
                   fill = "blue", 
                   alpha = 0.3) + 
          labs(title = colnames(d_clean1)[i], 
               x = "value", 
               y = "density") +
          theme(strip.text.x = element_text(size = 4), 
                axis.text.x = element_blank(), 
                axis.ticks.x=element_blank()) +
          annotate("text", -Inf, Inf, label = "Processed", hjust = 0, vjust = 1, color = "blue") + 
          annotate("text", -Inf, Inf, label = "Raw", hjust = 0, vjust = 3, color = "black"))
}
```

## 1.7 Unifying Data

### 1.7.1 Binding discrete and continuous data
```{r, include = TRUE, echo = FALSE, message = FALSE, warning = FALSE}
# PROCESSED DATA FRAME WITH DISCRETE ON LEFT AND CONTINUOUS ON RIGHT
new_flights <- cbind(d_clean1, c_clean)
```

### 1.7.2 Correcting additional dataset
```{r, include = TRUE, echo = FALSE, message = FALSE, warning = FALSE}
us_states2 <- us_states [,-3] %>% # removing "fips" column and dates after flights dataset
  mutate(date = as.Date(date)) %>%
  subset(date <= max(new_flights$FL_DATE))

length(which(rowSums(is.na(us_states2)) > 0)) # checking number of rows with NA values in data set
```

### 1.7.3 Joining datasets
```{r echo=FALSE, message=FALSE, warning=FALSE, include=TRUE}
# left join by date and ORIGIN STATE
fulldata <- left_join(new_flights, us_states2, by = c(c("FL_DATE" = "date"), c("ORIGIN_STATE_NM" = "state")))

# removing rows that have cases or deaths as "NA"
num_na_rows <- length(which(is.na(fulldata$cases) == TRUE | is.na(fulldata$deaths) == TRUE)) 
bad_rows <- which(is.na(fulldata$cases) == TRUE | is.na(fulldata$deaths) == TRUE)
fulldata2 <- fulldata[-bad_rows,]

# converting NA values to 0
fulldata3 <- fulldata
fulldata3[is.na(fulldata)] <- 0

#REMOVE LEADING HASH IN ONE OF THE LINES BELOW AND RUN, TO CONTRAST DIFFERENT PROCESSED DATAFRAMES
#fulldata # NA values are retained
#fulldata2 # NA values are omitted
#fulldata3 # NA values are converted to 0
```

## 1.7 Visualization (incomplete)
```{r, include = TRUE, echo = FALSE, message = FALSE, warning = FALSE}
#OPTIONAL
#for (i in 1:(ncol(c_clean1))) { # barplots of all continuous columns: processed
#  print(ggplot() + 
#          geom_density(aes(c_clean1[,i]), color = "blue", fill = "blue", alpha = 0.8) +
#          labs(title = colnames(c_clean1)[i], x = "value", y = "density")) 
#}

#OPTIONAL
#for (i in 1:(ncol(d_clean1))) { # barplots of all discrete columns: processed
#  print(ggplot() + 
#          geom_bar(aes(d_clean1[,i]), color = "blue", fill = "blue", alpha = 0.8) +
#          labs(title = colnames(d_clean1)[i], x = "value", y = "density")) 
#}
```

```{r, include = TRUE, echo = FALSE, message = FALSE, warning = FALSE}
ggplot(gather(c_clean1), aes(value)) + # processed continuous data density plots (small multiples)
  geom_density(color = "blue", fill = "blue", alpha = 0.8) + 
  facet_wrap(~ key, scales = "free") +
  theme(strip.text.x = element_text(size = 5), 
        axis.text.x = element_blank(), 
        axis.ticks.x = element_blank()) 

ggplot(gather(d_clean1), aes(value)) + # processed discrete data density plots (small multiples)
  geom_bar(color = "blue", fill = "blue", alpha = 0.8) + 
  facet_wrap(~ key, scales = "free") +
  theme(strip.text.x = element_text(size = 5), 
        axis.text.x = element_blank(), 
        axis.ticks.x=element_blank())
```

## Jonah's Section (Temporary Title)
Create smaller data frame to focus on the relationship between arrival delay with distance, origin, and (covid) deaths. 
```{r}
slim_data = fulldata3 %>% 
  select(ARR_DELAY, DISTANCE, ORIGIN, ORIGIN_STATE_NM, deaths)

names(slim_data) = tolower(names(slim_data))

#speed up initial plot development by randomly sampling 10% of the population
sample_row_index = sample(1:nrow(slim_data), nrow(slim_data)/10)
slim_sample = slim_data[sample_row_index,]
```

### Arrival Delay and Distance
The correlation between these variables is low, so I dont expect to find that distance will be helpful in predicting arrival delay
```{r}
no_delay = sum(slim_sample$arr_delay <= 0) / length(slim_sample$arr_delay) 
correlation = cor(slim_sample$arr_delay, slim_sample$distance)
```

Interesting pattern. It seems that most of the high arrival delays occur when the distance is short (< 1500 miles). Even given this, there does not appear to be a clear pattern that would suggest that these variables are closely related.
```{r}
ggplot(slim_sample, aes(distance, arr_delay)) +
  geom_point() +
  labs(x = "Distance (Miles)",
       y = "Arrival Delay (Minutes)",
       title = "Distance vs Arrival Delay")
```
This is the distribution of arrival delays with distance grouped in 1000 mile bins. Both high and low outliers have been filtered out. All distributions seem nearly identical. Notice that the third quartile is about zero for all distributions, meaning that most of the flights actually arrive early.
```{r}

#calculate quantiles to filter out outliers 
iqr =  quantile(slim_sample$arr_delay, 0.75) - quantile(slim_sample$arr_delay, 0.25)
upper_outlier_q = quantile(slim_sample$arr_delay, 0.75) + 1.5 * iqr
lower_outlier_q = quantile(slim_sample$arr_delay, 0.25) - 1.5 * iqr

slim_sample %>% 
  filter(arr_delay <= upper_outlier_q) %>% 
  filter(arr_delay >= lower_outlier_q) %>% 
ggplot(aes(distance, arr_delay)) +
  geom_boxplot(aes(group = cut_width(distance, 1000))) +
  labs(x = "Distance", 
       y = "Arrival Delay",
       title = "Arrival Delay Grouped in 1000 Mile Bins w/o Outliers")
```
On "short" flights, there appears to be no association between distance and arrival delay. Switching the equality sign revelas that flights longer than 1500 miles dont appear to have any clear pattern that would be useful in estimating arrival delay 
```{r}
slim_sample %>% 
  filter(distance < 1500) %>% 
ggplot(aes(distance, arr_delay)) +
  geom_point() +
  labs(x = "Distance (Miles)",
       y = "Arrival Delay (Minutes)",
       title = "Distance vs Arrival Delay (Distance < 1500 Miles)"
  )
  
```
In summary, I see explanatory relationship between distance and arrival delay

## Arrival Delay and Deaths
Weak correlation between deaths and arrival delay. 58% of days have no covid deaths. Scatterplot shows no hints of a relationship either
```{r}
cor(slim_sample$arr_delay, slim_sample$deaths)

zero_death_days = sum(slim_sample$deaths == 0) / length(slim_sample$deaths)

ggplot(slim_sample, aes(deaths, arr_delay)) +
  geom_point() +
  labs(x = "Covid Related Deaths",
       y = "Arrival Delay",
       title = "Covid Deaths vs Arrival Delay")
```
Fitting a linear model of arrival delay modeled by deaths shows that the coefficient of deaths is 0.0007. Clearly no predictive power. In summary, deaths is not found to be important in predicting arrival delays too
```{r}
lm = lm(arr_delay ~ deaths, data = slim_sample)
```

## Arrival Delay and Origin
Group by origin and find average delay for each. Then take the 10 that are the shortest delays on average, and the 10 that are the longest delays on average. Note there are 373 unique origins. 
```{r}
longest_origin_delays = slim_sample %>% 
  group_by(origin) %>% 
  summarize(mean_delay = mean(arr_delay)) %>% 
  arrange(desc(mean_delay)) %>% 
  head(10)

shortest_origin_delays = slim_sample %>% 
  group_by(origin) %>% 
  summarize(mean_delay = mean(arr_delay)) %>% 
  arrange(desc(mean_delay)) %>% 
  tail(10)
```

```{r}
ggplot(longest_origin_delays) + 
  geom_col(aes(rev(reorder(origin, mean_delay)), mean_delay)) +
  labs(x = "Origin", 
       y = "Mean Delay (Minutes)",
       title = "10 Worst Cities to Fly From")
```

```{r}
ggplot(shortest_origin_delays) + 
  geom_col(aes(reorder(origin, mean_delay), abs(mean_delay))) +
  labs(x = "Origin", 
       y = "Mean Minutes Arrived Early",
       title = "10 Best Cities to Fly From")
```
Fit a one way anova model to see if origin affects arrival delay. Value is essentially zero, meaning that at least one origin has an arrival delay mean that is different from the rest. Unsurprising that the null was rejected. Will explore this further in the refinement stage.
```{r}
model = aov(arr_delay ~ origin, data = slim_sample)
results = anova(model)
```
Same tables as above only for states instead of specific airports
```{r}
worst_states = slim_sample %>% 
  group_by(origin_state_nm) %>% 
  summarize(mean_delay = mean(arr_delay)) %>% 
  arrange(desc(mean_delay)) %>% 
  head(10)

best_states = slim_sample %>% 
  group_by(origin_state_nm) %>% 
  summarize(mean_delay = mean(arr_delay)) %>% 
  arrange(desc(mean_delay)) %>% 
  tail(10)
```

## Caitlyn's Section

Sampling 10% of total data and creating a tsibble to make time series plots

```{r}
slim_data2 = fulldata3 %>% 
  select(ARR_DELAY, FL_DATE, DEST_STATE_NM, cases, MKT_UNIQUE_CARRIER)
names(slim_data2) = tolower(names(slim_data2))

#speed up initial plot development by randomly sampling 10% of the population
sample_row_index = sample(1:nrow(slim_data2), nrow(slim_data2)/10)
slim_sample2 = slim_data2[sample_row_index,]

library("tsibble")
library("tsibbledata")
library("lubridate")
library("feasts")

group_cols <- c("fl_date", "dest_state_nm", "mkt_unique_carrier")

#group data by date, destination, and carrier and summarize mean delays and cases
group_slim_sample2 <- slim_sample2 %>%
  group_by(across(all_of(group_cols))) %>%
  summarise(delay_mean_on_given_day = mean(arr_delay),
            cases_mean_on_given_day = mean(cases))
group_slim_sample2 <- as.data.frame(group_slim_sample2) #ungroup for sibble

#tsibble indexed by date 
time_ts <- as_tsibble(group_slim_sample2, key = c("dest_state_nm", "mkt_unique_carrier"))

```

Visualizing time series for mean number of cases per day by destination state and carrier (midwest and western).

Visualizing time series for mean delay time per day by destination state and carrier (midwest and western). 

Thinking about adding column for US region (midwest, western, south, eastern) to dataset so that it is easy to facet visualizations by region or just so there aren't so many messy lines in plots. Implement shiny app for interactive time series plots?

```{r}
#filter to midwest destination states and dates after March that will have more cases
midwest <- time_ts %>%
  filter(dest_state_nm=="Illinois" | dest_state_nm=="Indiana" |
           dest_state_nm=="Iowa" | dest_state_nm=="Kansas" |
           dest_state_nm=="Michigan" | dest_state_nm=="Minnesota" |
           dest_state_nm=="Missouri" | dest_state_nm=="Nebraska" |
           dest_state_nm=="North Dakota" | dest_state_nm=="Ohio" |
           dest_state_nm=="South Dakota" | dest_state_nm=="Wisconsin")

western <- time_ts %>%
  filter(dest_state_nm=="Alaska" | dest_state_nm=="Arizona" |
           dest_state_nm=="California" | dest_state_nm=="Colorado" |
           dest_state_nm=="Hawaii" | dest_state_nm=="Idaho" |
           dest_state_nm=="Montana" | dest_state_nm=="Nevada" |
           dest_state_nm=="New Mexico" | dest_state_nm=="Oregon" |
           dest_state_nm=="Utah" | dest_state_nm=="Washington" |
           dest_state_nm=="Wyoming")

#mean cases per day by midwest destination state (after March 2020)
ggplot(midwest%>%filter(fl_date>="2020-03-01")) +
  geom_line(aes(x = fl_date, y = cases_mean_on_given_day, col=dest_state_nm)) +
  scale_fill_hue(l=100) +
  guides(color = guide_legend(override.aes = list(size = 3))) +
  labs(y="Mean cases per day", x="")

#mean delay time per day by selected midwest destination state
ggplot(midwest %>%filter(dest_state_nm=="Wisconsin"|dest_state_nm=="Illinois")) +
    geom_line(aes(x = fl_date, y = delay_mean_on_given_day, col=dest_state_nm)) +
  scale_fill_hue(l=100) +
  guides(color = guide_legend(override.aes = list(size = 3))) +
  labs(y="Mean delay time per day", x="")

#mean cases per day by western destination state
ggplot(western%>%filter(fl_date>="2020-03-01")) +
  geom_line(aes(x = fl_date, y = cases_mean_on_given_day, col=dest_state_nm)) +
  scale_fill_hue(l=100) +
  guides(color = guide_legend(override.aes = list(size = 3))) +
  labs(y="Mean cases per day", x="")

#mean delay time per day by western destination state (after March 2020)
ggplot(western) +
    geom_line(aes(x = fl_date, y = delay_mean_on_given_day, col=dest_state_nm)) +
  scale_fill_hue(l=100) +
  guides(color = guide_legend(override.aes = list(size = 3))) +
  labs(y="Mean delay time per day", x="")

```

Above time series for mean delays does not appear to indicate a significant difference before and during the pandemic.

Using K-S test to test whether distribution of mean delay times per day by destination state and carrier are significantly different pre-pandemic (before March 11, 2020) and during the pandemic (after March 11, 2020). Spoiler alert: delay times ARE significantly different (p<<0.05).

```{r}
pre_pandemic <- time_ts %>%
  filter(fl_date < "2020-03-11")
mean(pre_pandemic$delay_mean_on_given_day) #mean delay time pre-pandemic

pandemic <- time_ts %>%
  filter(fl_date >= "2020-03-11")
mean(pandemic$delay_mean_on_given_day) #mean delay time pandemic

#Kolmogorov-Smirnov test for difference in distributions of mean delays per day for pre-pandemic and pandemic flights
ks.test(pre_pandemic$delay_mean_on_given_day, pandemic$delay_mean_on_given_day)

```

Visualizing delays for different airline carriers.

```{r}
ggplot(time_ts) +
    geom_line(aes(x = fl_date, y = delay_mean_on_given_day, col=mkt_unique_carrier)) +
  scale_fill_hue(l=100) +
  guides(color = guide_legend(override.aes = list(size = 3))) +
  labs(y="Mean delay time per day", x="")

ggplot(time_ts) +
  geom_boxplot(aes(y=log(delay_mean_on_given_day), x=reorder(mkt_unique_carrier,delay_mean_on_given_day, sd),
               fill=mkt_unique_carrier)) +
  theme(legend.position = "none") +
  labs(y="Mean delay time per day (log scale)", x="Airline carrier")

```

Curious about whether the number of flights per day to each destination state changed after March 11th. SUspect that flights decrease dramatically after March. 

```{r}
#count number of flights per day for each destination state
number_flights <- slim_sample2 %>%
  group_by(dest_state_nm) %>%
  count(fl_date, sort=TRUE)

most_flights <- number_flights[0:10,] #CA has 9 of the top 10 number of flight days by state

number_flights_ts <- slim_sample2 %>%
  group_by(dest_state_nm) %>%
  count(fl_date, sort=TRUE) %>%
  as.data.frame() %>%
  as_tsibble(key="dest_state_nm")

#just looking at western states again since California has the most flights
western_no <- number_flights_ts %>%
  filter(dest_state_nm=="Alaska" | dest_state_nm=="Arizona" |
           dest_state_nm=="California" | dest_state_nm=="Colorado" |
           dest_state_nm=="Hawaii" | dest_state_nm=="Idaho" |
           dest_state_nm=="Montana" | dest_state_nm=="Nevada" |
           dest_state_nm=="New Mexico" | dest_state_nm=="Oregon" |
           dest_state_nm=="Utah" | dest_state_nm=="Washington" |
           dest_state_nm=="Wyoming")

ggplot(western_no) +
  geom_line(aes(x = fl_date, y = n, col=dest_state_nm)) +
  scale_fill_hue(l=100) +
  guides(color = guide_legend(override.aes = list(size = 3))) +
  labs(y="Number of flights", x="")
```


#Haoxuan's Section
```{r}
library(patchwork)
selected_data = fulldata3 %>% 
  select(ARR_DELAY, ORIGIN_STATE_NM, FL_DATE, AIR_TIME, cases)
```

#Plotting Date vs Arrival Delay with faciting.
#The delay has a decreasing trend over time. It has a higher delay time before April among all states on average.

```{r, fig.height = 20, fig.width = 25}
ggplot(selected_data) + 
  geom_line(aes(FL_DATE, ARR_DELAY, col = ORIGIN_STATE_NM)) + 
  facet_wrap(. ~ ORIGIN_STATE_NM) + 
  theme(legend.title = element_text(size = 10), 
        legend.text = element_text(size = 10)) + 
  theme_bw() + 
  labs(x = "Date", y = "Arrival Delay (mins)", title = "Delay Changes Over Days")
```

#Comparing Mean Cases and Mean Delay
#Calculate the mean cases and mean delay grouped by the origin state. It does not show that there are any correltation beteewn cases and mean delay.

```{r, fig.height = 20, fig.width = 25}
mean_cases = selected_data %>% 
  group_by(ORIGIN_STATE_NM) %>% 
  summarise(cases = mean(cases))

mean_delay = selected_data %>% 
  group_by(ORIGIN_STATE_NM) %>% 
  summarise(delay = mean(ARR_DELAY))

casesPlot = ggplot(mean_cases) + 
  geom_col(aes(cases, reorder(ORIGIN_STATE_NM, cases), fill = ORIGIN_STATE_NM)) + 
  scale_x_continuous(expand = c(0, 0)) + 
  theme_bw() + 
  theme(legend.position="none") + 
  labs(x = "Mean Covid Cases", y = "States")

delayPlot = ggplot(mean_delay) + 
  geom_col(aes(delay, reorder(ORIGIN_STATE_NM, delay), fill = ORIGIN_STATE_NM)) + 
  scale_x_continuous(expand = c(0, 0)) + 
  theme_bw() + 
  labs(x = "Mean Arrival Delay (mins)", y = "States")

casesPlot + delayPlot
```

#Plotting Cases vs Arrival Delay
# no clear pattern shown in the scatter plot, there are too many data points with no cases, and no clear pattern shown

```{r}
ggplot(selected_data) + 
  geom_point(aes(cases, ARR_DELAY))
             size = 0.5, alpha = 0.7) +
  theme_bw() +
  labs(x = "Covid Cases", 
       y = "Arrival Delay (mins)",
       title = "Covid Cases vs Arrival Delay")
```

#Plotting Air Time vs Arrival Delay.
#Flights with Airtime less than 400 minutes tend to have greater variability.
#Flights with Airtime more than 400 minutes never delay over 500 minutes.

```{r}
selected_data %>%
  ggplot() + 
  geom_point(aes(AIR_TIME, ARR_DELAY)
             size = 0.5, alpha = 0.7) +
  theme_bw() +
  labs(x = "Air Time (mins)", 
       y = "Arrival Delay (mins)",
       title = "Air Time vs Arrival Delay")
```
#Zooming in to flights that has less than 400 air time, arrivail delay has a decreasing trend as air time increases.
#After fitting a SLR model, the R-suqared value is small, it does not show linear relationship between air time and arrival delay.

```{r}
selected_data %>% 
  filter(AIR_TIME < 400) %>% 
  ggplot() + 
  geom_point(aes(AIR_TIME, ARR_DELAY), 
             size = 0.5, alpha = 0.7) +
  theme_bw() + 
  scale_x_continuous(expand = c(0, 0)) + 
  scale_y_continuous(expand = c(0, 0))
  labs(x = "Air Time (mins)", 
       y = "Arrival Delay (mins)",
       title = "Air Time (less than 400 mins) vs Arrival Delay")
  
airTime_model = lm(AIR_TIME ~ ARR_DELAY, data = selected_data)
summary(airTime_model)
```

